---
title: "Waterbird data cleaning and prep for analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

This serves as a worked example showing how to process ACR waterbird data from the raw database to a form usable for analysis. There are three major steps in this workflow:  
1. read the data, standardize field names and get fields in the correct format. Functions for these tasks are defined in waterbird_cleaning1_data_read_format.R  
2. separate birds field-identified to pooled species groups into constituent species. Functions for these tasks are defined in waterbird_cleaning2_split_groups.R  
3. handle any negative counts (birds that flew north past the boats); assign to the appropriate section and avoid double counting. Functions for these tasks are defined in waterbird_cleaning3_reconcile_negatives.R  

First load packages and setup
```{r}
library(tidyverse)
library(lubridate)
library(here)
library(RODBC)

library(birdnames)
custom_bird_list <- readRDS("C:/Users/scott.jennings/Documents/Projects/my_R_general/birdnames_support/data/custom_bird_list")


options("scipen"=999)
source(here("code/waterbird_cleaning1_data_read_format.r"))
source(here("code/waterbird_cleaning2_split_groups.r"))
source(here("code/waterbird_cleaning3_reconcile_negatives.r"))
```

These are the core waterbird taxa that are represented in this database. Other taxa exist in the db, but the following data management does not apply to them. 
```{r}
wbird_keep_taxa <- c("AMCOGRSCLESCBUFF", "AMCO", "COGA", "Anseriformes", "Alcidae", "Gaviidae", "Pelecanidae", "Podicipediformes", "Sterninae", "Suliformes")
#wbird_keep_taxa_gulls <- c("Anseriformes", "Alcidae", "Laridae", "Gaviidae", "Pelecanidae", "Podicipediformes", "Suliformes")
```


# Step 1. Read data and basic cleaning

Read the waterbirds data and do some basic data management in prep for splitting POOLED and reconciling negatives. Unless otherwise indicated, the functions are defined in waterbird_cleaning1_data_read_format.R.

read_species_table() and query_waterbirds() read data directly from the access database; the path where the database is stored must be specified for both functions, so we define it as db here for simpler code later. 
```{r}
db <- "C:/Users/scott.jennings/Documents/Projects/core_monitoring_research/water_birds/ACR_waterbird_data_management/data_files/waterbirds_v2.0.accdb"

```



```{r}
# read in just the species lookup table from the access db
spp_table <- read_species_table(db) 

# query 
wbirds <- query_waterbirds(db) %>% 
  clean_waterbirds() %>% # basic field names and formatting
  sppindex_to_alpha(spp_table = spp_table) %>% # the query sets some alpha.codes to the lookup number in the species table instead of the 4-letter code
  mutate(alpha.code = update_alpha(alpha.code)) %>% # from birdnames
  fix_precount_block_names() %>% # assign precounts to the correct section: bivalve, millertonbivalve and inverness into section 1; cypressgrove into section 2: walkercreek into section 3
  bird_taxa_filter(keep_taxa = wbird_keep_taxa) %>% # from birdnames; filter to just waterbird species
  separate(block, into = c("transect", "section"), remove = F) %>% 
  mutate(section = gsub("sec", "", section))  %>% 
  select(date, block, section, transect, alpha.code, count) # we only need a few columns

```

From here, you may wish to save wbirds if you want to use unprocessed data or want to do something with the data at the block level.  


# Step 2. handle pooled birds.

The logic of splitting is as follows:  

* There are several species of waterbirds that may not be identified to species but that can be assigned to a POOLED group (e.g. LOON, CORM).  

* POOLED birds are allocated to the constituent species of that POOLED group based on the ratios of positively identified birds in each constituent species. Allocation is only done if the number of positively identified birds in the group is greater than the number of undifferentiated birds.

* The bay is divided into 4 sections, numbered from south to north

* If there are enough identified birds at the section scale, then the ratio of IDed birds in each species counted in that section is used to divvy the lumped birds. The group cutoff value is 50 for grebes, 100 for all other groups.

* If there are not enough IDed birds in the Section, then the ratio for IDed birds across the entire Bay is used. NOTE: IT APPEARS KELLY AND TAPPAN 1998 DID NOT DO THIS EXTRA STEP. DOING THIS EXTRA STEP INCLUDES MORE INDIVIDUAL BIRDS, BUT AT THE COST OF MORE LIBERAL ASSUMPTIONS ABOUT RATIOS OF IDENTIFIED AND UNIDENTIFIED BIRDS BEING CONSTANT THROUGHOUT THE ENTIRE BAY. THERE ARE AT LEAST SOME OCCASIONS WHERE THE CONSTITUENT SPP RATIO IN THE SECTION WITH POOLED BIRDS IS QUITE DIFFERENT THAN THE RATIO IN THE WHOLE BAY (E.G., ) THE FOLLOWING CODE INDICATES WHETHER AN ALLOCATION WAS MADE BASED ON SECTION OR BAY RATIOS, SO THAT THE USER CAN DECIDE WHICH DATA TO INCLUDE IN ANALYSIS

Unless otherwise indicated, the functions are defined in waterbird_cleaning2_split_groups.R.  

For each pooled group on each day, we want to calculate the ratios of constituent species in that group (e.g. ratio of DCCO:BRAC:PECO on a given day that some birds were assigned to CORM). We need to do this at the section and bay scales.

The first part of step 2 is to sum the number of birds in each of the four sections of the bay. In the final mutate() line here we add a column "group.spp" for constituent species in each POOLED group, if applicable.  
sum_section_tallies_finals() incorporates birds counted in the precount zones into the appropriate sections.

```{r}
section_tallies_finals <- wbirds %>%  
  sum_section_tallies_finals() %>%
  mutate(group.spp = translate_bird_names(alpha.code, "alpha.code", "group.spp")) # from birdnames
```


Next we calculate section- and bay-scale ratios of constituent species.
* make_pooled_date_constituents() - for each date and section when birds are IDed to a POOLED group, this function extracts the number of birds IDed to the constituent species in that POOLED group on that day.
* calc_section_constituent_sum_proportion() - calculate the ratio of constituent species at the section level
* calc_bay_constituent_sum_proportion() - calculate the ratio of constituent species at the bay level


```{r}

constituent_ratios <- section_tallies_finals %>% 
  make_pooled_date_constituents() %>% 
  calc_section_constituent_sum_proportion() %>% 
  calc_bay_constituent_sum_proportion()
# warning about missing pieces is expected
```

And then we peel off just the POOLED bird data from wbirds, join with the ratio table, and allocate POOLED birds to constituent species based on the appropriate ratio.
```{r}

pooled_block <- get_pooled_block(wbirds)

allocated_pooled_block <- allocate_pooled_block(constituent_ratios, pooled_block) %>% 
  arrange(date, block, pooled.alpha.code, alpha.code)

```

A quick summary to view how many of each POOLED group were allocated by section of bay ratio or remain unallocated.  
```{r}

allocated_pooled_block %>% 
    filter(allocation.scale != "no.allocation") %>%  
    group_by(pooled.alpha.code, allocation.scale)%>% 
    replace(., is.na(.), 0) %>%
    summarise(total = sum(abs(allocated.count))) %>% 
  ungroup() %>% 
  pivot_wider(id_cols = pooled.alpha.code, values_from = total, names_from = allocation.scale) %>% 
  full_join(., allocated_pooled_block %>%
              filter(allocation.scale == "no.allocation") %>%
              distinct(date, section, pooled.alpha.code, pooled.section.tally) %>%
              group_by(pooled.alpha.code) %>%
              summarise(no.allocation = sum(abs(pooled.section.tally))) %>%
              ungroup()) %>% 
  select(pooled.alpha.code, section, bay, no.allocation) %>% 
  view()

```

The final step in the splitting sub-workflow is to combine the split pooled, unsplit pooled, and unpooled birds back together.

```{r}

#split_unsplit_unpooled <- combine_split_unsplit_unpooled(section_tallies_finals, allocated_pooled)
split_unsplit_unpooled_block <- combine_split_unsplit_unpooled_block(wbirds, allocated_pooled_block)

```


# Step 3. Handle negatives  

## The assumptions and logic for handling negatives:  

* This method of handling negatives relies on data entered at the block scale. Each block with either have a net negative or net positive number of birds. We assume that birds counted as positive in one block are not counted as negative in another block in that section, and birds counted as negative in one block are not counted as positive in another block in that section. This might be the most tenuous assumption of this process, and when violated it could lead us to overestimate the total number of birds on the bay (double counts birds encountered twice in the same section). But it is also the simplest assumption we can make for this scenario. We don't currently have information to tease apart movements between blocks within a section, and we will need to make other tenuous assumptions if we want to try accounting for this lateral movement with the data as it is currently entered (i.e. assume how many times a bird/flock might be re-encountered within a section before it either flies forward to the next section or stays permanently behind the boats). If we had each individual tally in each block entered digitally then we could possibly track whether similar sized flocks were counted in multiple blocks in the same section.   

* We assume all birds flying past the boats originated on the bay (did not arrive to the bay behind the boats from the south during the count). Thus, all birds flying past the boats have either already been counted or they were missed in an area that was counted. 

* At the end of each section, there will be known birds behind the boats (positive counts) and known birds ahead of the boats (negative counts). 
  + All of these birds have been counted at least 1 time. 
  + Some of them have been counted 2 or more times. 
  
* We want to assign birds to the section where they were counted the first time.

* All positives should be added to the section where they were observed unless their presence in that section could be the result of net negatives from the prior sections.

* All negatives should be assigned to a section before the one they were detected in (and they should never be added to section 4), and because of that they should be subtracted from future section counts. 
  + If the negatives can be explained by positives in all previous sections they have already been assigned to the section they were first observed in and they should be subtracted from the current and future sections. 
  + If they cannot be explained by previous positives they should be added back to section 1 (most parsimonious assumption is that they were missed in precount or moved south between end of precount and start of boat count).
  
This final assumption leads us to a conservative estimate of the number of birds on the bay. Since we assume any birds flying north past us are re-encountered as soon as we pass another positive bird, we are estimating the minimum number of birds known to be on the bay that day. This key assumption may be false, so we may be underestimating the number of birds on the bay. But we stick with this assumption because we don't have a feasible way to test it and because we would rather underestimate than overestimate the number of birds on the bay.

## The mechanics of handling negatives:  

At the end of each section, we tally up the known birds behind and ahead of the boats in that section. We compare those tallies to the cumulative birds ahead of and behind the boats at the start of that section to determine how many new birds we actually observed in that section.  

* Any new observed negatives in the current section are added to the cumulative negatives.  

* Any new observed positives are added to the cumulative positives.  

* If we started the section with any cumulative negatives, then we assume those birds were re-encountered in this section and we subtract those cumulative negatives from the current section positive tally. 

  + If there are more current section positives than cumulative negatives, then the number of new positives in the current section is the number of observed positives minus the cumulative negatives.  

  + If there are more cumulative negatives than current section positives, then we add no new positives for the current section and the balance of cumulative negatives (once current section positives have been subtracted) is carried forward to the next section.  

* If there are more cumulative negatives than cumulative positives, then that difference represents birds that we missed somewhere along the way; the most logical assumption is that we missed them between the end of the precount and the start of the boat count, so we add them back to section 1.  


Once we do this for all 4 sections, we have:  

* a value for known new birds added to each section  

* a value for number of missed birds to add back to section 1  

* a value for the number of known birds ahead of the boat  

* a value for the number of known birds behind the boat  


As a check on the process, the sum of the known birds ahead of and behind the boats at the end of section 4 should equal the sum of the number of birds assigned to each section plus those to be added back to section 1. This test is formalized below with check_subtracted_forward().


A quick look at how many dates we have negatives for each species in each block:

```{r}
filter(wbirds, count < -100) %>% 
  count(alpha.code, block) %>% 
  pivot_wider(id_cols = alpha.code, values_from = n, names_from = block) %>% 
  select(alpha.code, contains("1"), contains("2"), contains("3"), contains("4")) %>% 
  view()

```

And now the series of functions to do the negative handling.
```{r}

subtracted_forward <- split_unsplit_unpooled_block %>% 
  get_section_pos_neg() %>% 
  widen_block_pos_neg() %>% 
  subtract_forward_add_back()

```

subtracted_added_viewer() is a function that reshapes a single species X date record for easier viewing (and hopefully understanding) of how negatives were handled. See the included new_negative_logic_examples.pdf for some worked examples of how negatives are handled. The examples are for the following species dates: BRAC 2004-12-18, 2009-02-08, 2013-02-09, 2015-02-14; DCCO 2019-01-12
```{r}

subtracted_added_viewer(zdate = "2004-12-18", zalpha.code = "BRAC")

subtracted_added_viewer(zdate = "2009-02-08", zalpha.code = "BRAC")

subtracted_added_viewer(zdate = "2019-01-12", zalpha.code = "DCCO")

subtracted_added_viewer(zdate = "2013-02-09", zalpha.code = "BRAC")

subtracted_added_viewer(zdate = "2015-02-14", zalpha.code = "BRAC")


```


Check for any records where the sum of the known birds ahead of and behind the boats at the end of section 4 does NOT equal the sum of the number of birds assigned to each section plus those to be added back to section 1.
```{r}

check_subtracted_forward(subtracted_forward) %>% 
  arrange(-diff.total) %>% # all diff.total should be 0, this puts any >0 at the top; 
  view()


```

Finally, summarize the negative handled data (mostly just adding the net negatives back to section 1). waterbirds4_analysis is the data ready for any analysis, data visualization, etc.
```{r}

waterbirds4_analysis <- make_wbirds4_analysis(subtracted_forward)

saveRDS(waterbirds4_analysis, "data_files/working_rds/waterbirds4_analysis")

```

